KIN: Method to infer relatedness using low coverage ancient DNA

### 2022-04-05 paper_overleaf/main.tex

Identifying related individuals is a common task in genetic studies. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential  step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population.

In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity. ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}

@@

However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS \cite{vai_KINship_2020, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}. One exception is SNPduo, a software that uses IBS directly from a pair of individuals to visualize and analyze their relatedness \cite{roberson_visualization_2009}.

\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings without and with ROH. In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases. p can take values of $p_0$, $p_1$ or $p_2$ which are expected proportion of differences in IBD states 0,1 and 2 respectively.  }
    \label{fig0:schematic}
\end{figure}

@@

In cases where the allele frequency or a reference panel for the target population is not available, it is possible to use allele frequency from a modern population from the same geographic location, or to try using population allele frequencies from multiple potentially close populations. However, incorrect assumptions may cause the individuals in the target population to look more similar or different to each other \cite{amorim_understanding_2018}. Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population. It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen. In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding. Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.

@@

\subsection{Contamination correction}\label{contam}
Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}. Even levels of contamination may make samples look less similar, and thus distort relatedness analyses. This issue is particularly pronounced  when analyzing populations with low diversity, where even $<5\%$ contamination can cause loss of power to detect family relations . To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.

We conceptualize contamination in the following way: If data from individual $i$ has a known contamination rate of $C_i$, this means that any particular read  will actually come from the desired individual with probability $1-C_i$. With probability $C_i$ the read comes from a contaminant. Thus, when comparing two reads, one drawn from individual $i$ and the other drawn from individual $j$, there are three possible cases: First, with probability $(1-C_i)(1-C_j)$, both reads are endogenous and we can use them to infer relatedness. With probability $C_i(1-C_j) + C_j(1-C_i)$, one read is endogenous and the other is contaminant. Finally, with probability $C_iC_j$, both reads are contaminant. If we make the simplifying assumption that both $C_i$ and $C_j$ are small, the $C_iC_j$ terms  can be ignored. Then, we compare contaminant read to endogenous read with the probability $C_{ij} = C_i + C_j$.

We assume that the level of contamination is known (cite Stephane and Kay's review). We  estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$. To estimate $\mathbb{E}[D_w^{'}]$, we first need to calculate $P(c=0|d=1)$. Here, $c$ is a binary variable with value $0$ or $1$ denoting that the comparison at a site is between endogenous reads, or between endogenous and contaminant reads respectively. Similarly $d$ is a binary variable taking values 0 and 1 for seeing no difference, or a difference, when comparing the reads at a given site. We can further expand this probability:

\begin{align}\label{eq:12a}
    P(c=0|d=1) &= \frac{P(d=1|c=0)P(c=0)}{P(d=1)} \nonumber\\
\end{align}

Here, $P(d=1|c=0)$, referred later to as $\rho_{ij}$ for simplicity, is equivalent to the proportion of differences when we compare endogenous reads from individuals $i$ and $j$. This can be estimated by considering that the proportion of pairwise differences $P(d=1) =\frac{\sum_w D_w}{\sum_w N_w}$, referred to as $o_{ij}$ later, can be expressed as a linear combination of the proportion of differences between endogenous reads and that between endogenous and contaminant reads \cite{peyregne_present-day_2020}. 
\begin{align}\label{eq:13}
    o_{i,j} = (1-C_{ij}) \rho_{ij} + C_{ij} \phi \nonumber\\
    or, \rho_{ij} = \frac{o_{ij} - C_{ij} \phi} {1 - C_{ij}}
\end{align}

Here, $\phi$ is the probability of seeing a difference when an endogenous read is compared to a contaminant read, and is equivalent to the average sequence divergence between individuals of the target and contaminating populations.
Given this information, we can rewrite eq. \ref{eq:12a}:
\begin{align}\label{eq:12}
    P(c=0|d=1) &= \frac{(1 - C_{ij}) \rho_{ij}}{(1 - C_{ij}) \rho_{ij} + C_{ij} \phi} \nonumber\\
\end{align}
We can further write $\mathbb{E}[D_w^{'}]$ as $D_w P(c=0|d=1)$:
\begin{align}
    \mathbb{E}[D_w^{'}] &= D_w\frac{(1-C_{ij})\rho_{ij}}{(1-C_{ij}) \rho_{ij} + C_{ij} \phi}\\
\end{align}

Similarly, we derive $\mathbb{E}[S_w^{'}]$ as the estimate for the number of sites that do not show pairwise differences.
\begin{align}
    \mathbb{E}[S_w^{'}] &= (N_w-D_w) P(e=1|d=0) = (N_w - D_w)\frac{(1-\rho_{ij})(1-C_{ij})}{(1-\rho_{ij})(1-C_{ij}) + C_{ij}(1-\phi)}\\
\end{align}

%adding back the previous versions:
If we look at one site in the window, we could find either no difference in reads from individuals $i$ and $j$ ($d_s$=0), or a non-zero difference ($0<d_s<=1$). Here $d_s$ is the probability of seeing a difference at site $s$, and $\sum_{s \in w} d_s =D_w$. We ask, what is the

If we make the simplifying assumption that both $C_i$ and $C_j$ are small, terms of the order of $C^2$ can be ignored and there are just two cases: with probability $C_{i,j} =C_i + C_j$, one of the reads stems from a contaminant, and we effectively compare a contaminant fragment with an endogenous one. The probability that they differ is the average divergence between target and contaminating population, which we denote with $\phi$. The other case (w. p. $(1-C_{i,j}$) is where we compare two endogenous reads, which will differ with probability

\begin{equation}
    p_{e} = \frac{p_{obs}-\phi C_{i,j}}{1-C_{i,j}}
\end{equation}

Here, $$p_{obs} = \frac{\sum_w D_w}{\sum_w N_w}$$

We estimate $\hat{D_c} = C_{i,j} N_{tot} \phi$ and $\hat{D_e} = (1-C_{i,j}) N_{tot} p_{obs}\phi$.
If there are $N_{obs}$ overlapping sites in a window, $N_{c} = C_{i,j}N_{obs}$ of them will include a contaminant allele, and $N_e = (1-C_{i,j})N_{obs}$ will be between endogenous alleles.

For KIN, we can use $N_e$ as an input (for each window), but we also need the proportion of differences, $D_e$. For this, we use the assumption that the proportion of sites where the contaminant differs from our samples is

$$\phi = \frac{D_c}{N_c} = \frac{D_c}{C_{i,j} N_{obs}}, $$
and thus $D_c = C_{i,j} \phi N_{obs}$.

We can think of $p_c$ and $\rho$ as empirical likelihoods that a particular site shows a difference between two reads, given that the comparisons contain only endogenous or one contaminant read, respectively. 
\begin{align}
    P(E|D) &= \frac{P(D|E)P(E)}{P(D)} \nonumber\\
    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs}\frac{p_e (1-c)}{p_e(1-c) + c\rho}\\
    \mathbb{E}[S_e] &= D_{obs} P(E|\text{not} D) = S_{obs}\frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c(1-\rho)}
\end{align}

And for calculating $N_{e}$, we added $D_{e}$ to $S_{e}$

The goal is that given $C_{i,j}, p_e, \BN and \BD$ to calculate adjusted $\hat{\BN}$ and $\hat{\BD}$ that reflect the endogenous proportion only.

, and the observed proportion of differences between individuals $i$ and $j$ as $p_{oij}$. We calculate average endogenous pairwise difference $p_{eij}$ for individuals i and j as follows \cite{noauthor_ancient_nodate}:
\begin{align}
    p_{oij} = C_{ij}  p_c + (1-C_{ij})  p_{eij}\\
    or, p_{eij} = \frac{p_{oij} - C_{ij}  p_c}{1-C_{ij}} 
\end{align}

Given $p_{eij}$, we would ideally like to calculate the complete data log likelihood for the relatedness model while summing over all possible differences we could have in a window weighted by the probability of seeing that difference.

\paragraph{Short version}
With contamination, we do not observe the number of differences $D_{true}$ and total number of comparisons $N_{true}$ directly, but instead need to approximate them. The window-subscript is omitted in the following for clarity:
\begin{equation}
    P(D_{\text{obs}} | N_\text{obs}, Z, \theta) \approx 
    P( \hat{D} | \hat{N}, Z, \theta)
\end{equation}
We do that by calculating the probability that a single read is endogenous given it has a difference.
\begin{align}
    \hat{D} = \mathbb{E}[D | D_{\text{obs}}, c] &= 
    D_{\text{obs}} P(E_i | D_i=1) \\
    &= D_\text{obs} \frac{P(D_i=1 | E_i)P(E_i)}{P(D_i=1)}\\
    &= D_\text{obs} \frac{(1-c)p_e}{(1-c)p_e + c p_c}
\end{align}

and likewise for reads that do not have a difference
\begin{align}
    \hat{S} = \mathbb{E}[S | S_{\text{obs}}, c] &= 
     S_\text{obs} \frac{(1-c)(1-p_e)}{(1-c)(1-p_e) + c (1-p_c)}
\end{align}

and $\hat{N} = \hat{S} + \hat{D}$

\begin{equation}
    P(D_{\text{obs}} | Z, N_\text{obs}, \theta) = \sum_n\sum_d P(D_{\text{obs}}, N_\text{obs} | D_w=d, N_w, C_i, C_j)\underbrace{P(D_w=d | N_w, Z, \theta)}_\text{old likelihood}
\end{equation}

i.e. we sum over all possible $D_w$, $N_w$. Instead, we use

\begin{align}
    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
    &= [\prod_{w} [\sum_\kappa P(D_{cor,w}|Z_w, H_w, \theta,c,p_c) P(D_{cor,w}=\kappa)] \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
\end{align}

We realize that this calculation would be time consuming, and would further complicate the HMM. To avoid this, instead of using weighted sum of all possible $\kappa$, we use $E(D_{cor,w}|Z_w, H_w, \theta,c,p_c)$. This simplifies the equation 21 to :

\begin{align}
    P(D_{cor},Z|D,H,\theta,c,p_c) &= P(D_{cor}|Z,D,H,\theta,c,p_c) P(Z|D,H,\theta,c,p_c)\nonumber\\
    &= [\prod_{w} P(E(D_{cor,w})|Z_w, H_w, \theta,c,p_c) \prod_{w} P(Z_w|Z_{w-1}, \theta)] P(Z_0| \theta)
\end{align}

We calculate $E(D_{cor,w})$ as shown:

\begin{align}
    E[N_e] + E[N_c] = N\\
    E[D_e] + E[D_c] = D
\end{align}

Probability of comparing endogenous reads at a site where we see a difference:
$$P(\xi | N = 1,D = 1)=\frac{(1-c) p_e}{c p_c + (1-c) p_e} $$
Probability of comparing endogenous reads at a site where we see no difference:
$$P(\xi | N = 1,D = 0)=\frac{(1-c)(1-p_e)}{c (1-p_c) + (1-c) (1-p_e)} $$
Expectation of number of endogenous comparisons in a window:
\begin{align}
    E[\xi | N, D] = D E[\xi | N=1, D=1] + (N-D) E[\xi | N=1, D=0]\\
    = D P(\xi | N=1, D=1) + (N-D) P(\xi | N=1, D=0)
\end{align}

Total number of endogenous sites showing a difference in a window: 
$$D_{cor,w} = D_w* \frac{p_e(1-c)}{p_e(1-c) + c p_c} $$
Total number of endogenous sites showing 0 difference in a window: 
$$S_{cor,w} = S_w* \frac{(1-p_e)(1-c)}{(1-p_e)(1-c) + c (1-p_c)} $$
Total number of endogenous sites in a window:
$$N_{cor,w} = D_{cor,w} + S_{cor,w}$$

This model outputs the contamination-corrected number of differences and overlapping sites in each window.

We added $D_w^{'}$ to $S_w^{'}$ to get the corrected number of overlapping sites $N_w^{'}$

@@

\paragraph{Comparing KIN-HMM with READ}
In Figure Fig. \ref{fig4:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios. We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  except for $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages. We show that adding contamination adversely affects both the False Positve and True Positive rates corresponding to 1st and 2nd Degree relations for READ, while this does not affect KIN performance too much.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/comparison_plot.png}
    \centering
    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH. Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated. 'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
    \label{fig4:Comparison_READ_KIN}
\end{figure}

\paragraph{Evaluating IBD prediction}
KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child. We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals. We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}). Fig. shows that we can predict IBD states for Unrelated $3^{rd}$ Degree, $2^nd Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling. On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
    \centering
    \caption{Evaluation of IBD estimation at different coverages. y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean.}
    \label{fig4:Comparison_READ_KIN}
\end{figure}

\subsubsection{Application to Chagyrskaya Neanderthals}
To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia. This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya . The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena (again cite Laurits paper). All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}. is dataset has all the issues that KIN deals with, and hence we decided to apply it here to see if we could find family relationships that would not be possible with other available methods. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). We also find that Chagyrskaya 06 and Chagyrskaya 14 are classified as parent-child with low confidence, but from the morphology they likely stem from the same individual. We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively). Further, we identify   Chagyrskaya07 and Chagyrskaya60 as 3rd-degree relatives.

@@

We compared our IBD estimates to lcMLKIN for all pairs for whom READ results (with s.d. >1) and KIN results ($\Delta LL$>1) match (fig. S\ref{figS6:Chagyrskaya_ibd}).  We find that lcMLKIN does create four different clusters for different relatedness. However, the clusters are strongly deviated from the expected values, likely due to the ROH.  Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $r \approx 0.6, k_0 \approx 0.33$.

@@

We further applied KIN to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019. We compared our relatedness estimates to those obtained from the authors using READ and lcMLKIN, and found that in most cases the three methods agree. Out of 6903 pairwise comparisons, 346 pairs show different relatedness for KIN and READ (table S2). However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Lambda >1$. When we investigated some of these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig. S\ref{figS8:eg1}). Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLKIN (more than 10,000 overlapping SNPs). Among these 20 pairs, we found that lcMLKIN matches KIN's estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLKIN predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLKIN predict unrelated, while KIN estimates 3rd degree with log likelihood=1.7).  
We further found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3). There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $loglikelihood ratio>2$ for KIN. Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and again the heterozygosities along the genome match the inference of KIN (fig. S\ref{figS9:eg2}).

\section{Discussion}\label{discussion}

Here, we present a new method called KIN to estimate genetic kinship and the location IBD tracts from low coverage data, taking into account long ROH, contamination, as well as ascertainment bias. Our method utilizes a set of HMMs to estimate IBD tracts, conditional on a particular relatedness, and we use the log likelihoods of these HMMs to find the best fit.

We used simulated data to validate KIN's performance. We compared KIN's model performance to a widely used software READ to show that our approach gives similar predictions as READ on ideal data with 4x coverage, and without long ROH, ascertainment and contamination, while additionally distinguishing between parent-child and siblings, and identifying third degree relatives. We demonstrate that at low coverage (0.1x) we have higher power to detect first degree relatives compared to READ with similar false positives, while a higher power to identify second degree relatives at the cost of slightly higher false positive rate.

Counter-intuitively, we see that adding long ROH ($~17\%$) to simulations improves the performance of both KIN and READ. READ's performance improves because ROH does not cause any change in the average pairwise differences between unrelated individuals, while decreasing pairwise differences between related individuals. The higher the relatedness, the higher is the decrease in pairwise differences in presence of ROH. And so, ROH causes higher power for READ which otherwise is a conservative method. KIN performs well for the same reasons, but distinguishing IBD and ROH becomes important for distinguishing different relationship cases within a degree, i.e. when distinguishing between parent-child and siblings.

Contamination, on the other hand, can affect the performance of READ adversely depending on the amount of exogenous DNA, and the number of individuals contaminated. We show that even at contamination levels  $\leq 3\%$, and contamination in only 8 out of 17 individuals, READ performance falls. Contamination in more individuals causes a shift in median difference in the population and may cause non-contaminated individuals to look more related, thus increasing the false positive rates. On the other hand, an increase in the amount of contamination in a few individuals can make them look less related to uncontaminated individuals. Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compare to naive methods such as READ, although they too could be ameliorated in a similar way (cite Laurits). SNP ascertainment causes a decrease in power for both  methods, since it reduces the amount of data available. However, the decrease in power is lower for KIN compared to READ.

>>>>

Identifying related individuals is a common task in genetic studies. Relatedness is of direct interest in e.g. DNA forensics, where familial search can aid in solving criminal cases, and to identify unknown deceased persons \cite{murphy_law_2018,ram_genealogy_2018}. Genetic paternity tests have an important application in resolving family relation, e.g. in establishing relationship between a person applying for immigration and the claimed relatives \cite{egeland_beyond_2000}. It is also an essential  step in population genetics and association studies, where samples are typically assumed to be independent random draws from the population. For animal and plant breeders and conservation biologists, reconstructing pedigrees and finding related individuals is important to avoid inbreeding and ensure diversity ~\cite{habier_impact_2007,oliehoek_estimating_2006,kardos_measuring_2015}.

In ancient DNA studies, relatedness can be used to identify bones and teeth belonging to the same individual, and can provide an understanding of an ancient society's organization and hierarchy, social structures, and cultural aspects ~\cite{baca_ancient_2012,mittnik_KINship-based_2019,sikora_ancient_2017}.

@@

However, it is not possible to directly observe IBD segments, and so a common approach is to first identify segments of the genome that are Identical by State (IBS) and using population allele frequencies obtained from a reference panel to calculate the probability of IBD given IBS \cite{vai_KINship_2020, boehnke_accurate_1997,lynch_estimation_1999, albrechtsen_natural_2010, purcell_plink_2007}.
\begin{figure}[h!]
    \includegraphics[width=18cm]{plots/inkscape_finalImg/schematic_sib.png}
    \centering
    \caption{IBD sharing between siblings without and with ROH. In top row, recombined chromosomes inherited from parents by a pair of siblings are shown. On the left, the parents have unrelated chromosomes, and on the right one parent has a homozygous pair of chromosomes. The bottom row shows the expected proportion of differences (p) along the chromosome in both the cases. p can take values of $p_0$, $p_1$ or $p_2$ which are expected proportion of differences in IBD states 0,1 and 2 respectively.  }
    \label{fig0:schematic}
\end{figure}

@@

Several methods have been proposed to estimate relatedness without a reference panel, but many of these require either $>4$x coverage \cite{waples_allele_2019}, or a large sample size to get an estimate of population allele frequencies from the samples \cite{theunert_joint_2017}. In this study, we show that contamination can cause some of the individuals in ancient populations to look more unrelated than they are. This happens because the DNA sequence of an ancient individual typically exhibits more differences to a contaminant sequence than that of another ancient individual from the same population. It is worth mentioning that different DNA libraries may have different levels of contamination, even when they have been sampled from the same specimen. In addition, some analyzed  genomes  may have long runs of homozygosity (ROH), for example due to a small population size, or due to recent inbreeding. Long ROH cause related individuals to seem genetically more similar to each other, but do not affect the genetic distance between unrelated individuals.

@@

\subsection{Contamination correction}\label{contam}
Contamination by the DNA from present-day people is a common feature in human ancient DNA \cite{peyregne_present-day_2020}. To address this issue, we develop a heuristic that adjusts both $D_w$ and $N_w$ to remove contamination from the data.

We assume that contamination rates in both individuals are known and small, and set $C_{ij} \approx C_i + C_j$, where $C_i, C_j$ are the contamination estimates from the two individuals. We also assume the divergence $\phi$ between our target population and the putative contaminant is known. With probability $C_{ij}$, a comparison between two random reads from either individual will contain a contaminant read, and thus contain a difference with probability $\phi$, and with probability $1 - C_{ij}$ it will be between endogenous ones. The comparisons between two contaminant reads are ignored, which is why we assume $C_{ij}$ to be small.

We  estimate the expected number of differences from comparison of endogenous reads $\mathbb{E}[D_w^{'}]$, and the total number of sites with overlapping endogenous reads $\mathbb{E}[N_w^{'}]$.

For any particular comparison showing a difference $D$, we calculate the probability of the event $E$ that it is between endogenous reads as 
\begin{equation}
    P(E|D) = \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}.
\end{equation}

Then, by linearity of expectation, we obtain our estimators for the expected number of endogenous comparisons with a difference as 
\begin{align}
    \mathbb{E}[D_e] &= D_{obs} P(E|D) = D_{obs} \times \frac{P(D|E)P(E)}{P(D|E)P(E) + P(D| \neg E)P( \neg E)}
\end{align}.

Of these terms, $P(E)= 1 -C_{ij} = 1 - P(\neg E)$, and $P(D| \neg E) = \phi$.

For $P(D|E)$, we use an estimator based on the genome-wide average:

\begin{align}
    P(D|E) = p_e = \frac{\frac{\sum_w D_w}{\sum_w N_w} - C_{ij} \phi}{1 - C_{ij}}.
\end{align}

Taken together,

\begin{equation}
 \mathbb{E}[D_e]= D_{obs}\frac{p_e (1-C_{ij})}{p_e(1-C_{ij}) + C_{ij}\phi}
 \end{equation}

 Analogous considerations lead to the expected number of endogenous comparisons that yield no difference:

 \begin{equation}
\mathbb{E}[S_e] = S_{obs} P(E| \neg D) = S_{obs} \frac{(1-p_e)(1-C_{ij})}{(1-p_e)(1-C_{ij}) + C_{ij}(1-\phi)}.
 \end{equation}

Finally, we set $\mathbb{E}[N_e] = \mathbb{E}[S_e] + \mathbb{E}[D_e]$

@@

\paragraph{Comparing KIN-HMM with READ}
In Figure Fig. \ref{fig4:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios. We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  except for $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages. We show that adding contamination adversely affects the True Positive rate for $2^{nd}$ Degree, and both the False Positve and True Positive rates corresponding to 2nd Degree relations for READ, while this does not affect KIN performance too much.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/comparison_plot.png}
    \centering
    \caption{Comparison of KIN with READ using simulations with different coverages, and different cases of ascertainment, contamination and ROH. Unrelated label here refers to KIN performance results when all Unrelated, Fifth Degree, Fourth Degree, Third Degree pairs are labelled as Unrelated. 'Unrelated w/o $3^{rd}$ Degree' refers to the performance results when $3^{rd}$ Degree is classified separately from the unrelated individuals.}
    \label{fig3:Comparison_READ_KIN}
\end{figure}

\paragraph{Evaluating IBD prediction}
KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child. We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals. We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}). Fig. \ref{fig4:IBDstate_accu} shows that we can predict IBD states for Unrelated $3^{rd}$ Degree, $2^nd Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling. On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.

\begin{figure}[h!]
    \includegraphics[width=16cm]{plots/plotimg/plot_IBDaccuracy.png}
    \centering
    \caption{Evaluation of IBD estimation at different coverages. y-axis shows accuracy calculated over 60 simulations with each relatedness case and six different coverages shown on x-axis. The error bars are drawn at 1 standard deviation from the mean.}
    \label{fig4:IBDstate_accuracy}
\end{figure}

\subsubsection{Application to Chagyrskaya Neanderthals}
To test KIN on real ancient dataset, we applied it to a Neandertal data set  from Chagyrskaya cave in Siberia \cite{kolobova_archaeological_2020-1}, Russia. This dataset contains genetic data from  14 fossils that might  possibly belong to contemporary Neandertals who occupied the cave 63 ± 4 to 48 ± 3 kya (Laurits et al., 2022, in press). The data has low-to intermediate coverage ranging from 0.01x to 12.34x, with 8 samples at  $<1x$ coverage. Some of these specimens showed signs of long ROH, and contamination from modern humans as well as Hyena (Laurits et al., 2022, in press). All these samples were captured with an array targeting for variable sites in high coverage Neandertal and Denisovan genomes and common variations in Africans (Laurits et al.,2022), and we further restricted  this data to the variable sites in two high-coverage Neandertal genomes: Altai Neandertal (Denisova 5) \cite{prufer_complete_2014} and Vindija 33.19 \cite{prufer_high-coverage_2017}. Our results for pairwise relatedness for these individuals is shown in Fig.\ref{fig5:Chagyrskaya_KIN}. We found three fossils from the same  individual (Chagyrskaya13-Chagyrskaya19-Chagyrskaya1141), a parent-child pair  (Chagyrskaya07 and Chagyrskaya17), and a pair of  $2^{nd}$-degree relatives (Chagyrskaya01/Chagyrskaya60). We also find that Chagyrskaya 06 and Chagyrskaya 14 are classified as parent-child with low confidence, but from the morphology they likely stem from the same individual. We believe that the contradiction is due to non human contamination present in these libraries ($1\%$ and $2.9\%$ respectively). Further, we identify   Chagyrskaya 07 and Chagyrskaya 60 as 3rd-degree relatives.

@@

We compared our IBD estimates to lcMLKIN for all pairs for whom READ results (with s.d. $>$1) and KIN results ($\Delta LL>$1) match (fig. S\ref{figS6:Chagyrskaya_ibd}).  We find that lcMLKIN does create four different clusters for different relatedness. However, the clusters are strongly deviated from the expected values, likely due to the low coverage and ROH.  Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $r \approx 0.6, k_0 \approx 0.33$.

@@

We further applied KIN to a genome-wide dataset of 118 ancient individuals sequenced and analyzed by Mittnik et al.,2019 \cite{mittnik_kinship-based_2019}. We compared our relatedness estimates to those obtained from the authors using READ and lcMLKIN, and found that in most cases the three methods agree. Out of 6903 pairwise comparisons, 346 pairs show different relatedness for KIN and READ (table S2). However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Lambda >1$. When we investigated some of these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig. S\ref{figS8:eg1}). Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLKIN (more than 10,000 overlapping SNPs). Among these 20 pairs, we found that lcMLKIN matches KIN's estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLKIN predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLKIN predict unrelated, while KIN estimates 3rd degree with log likelihood=1.7).  
We further found 43 pairs for which READ estimates first degree, and compared the estimates of lcMLKIN and KIN for these cases (suppl table3). There were a total of 20 pairs where lcMLKIN had sufficient data, and 26 pairs where $\Delta LL>1$ for KIN. Among the 19 overlapping pairs, we found four pairs where KIN and lcMLKIN differ, and again the heterozygosities along the genome match the inference of KIN (fig. S\ref{figS9:eg2}).

\section{Discussion}\label{discussion}
Here, we present a new method called KIN to estimate genetic kinship and the location of IBD tracts from low- coverage data, taking into account long ROH, contamination, as well as ascertainment bias. Our method utilizes a set of HMMs to estimate IBD tracts and uses them to classify each pair of individuals into a possible relatedness case, along with a measure of classification confidence. We evaluated the method performance of KIN, and compared it to that of READ using simulated pedigrees. Finally, we show applications of KIN on two ancient datasets.

In our simulations, we find that adding long ROH ($~17\%$) to simulations improves the performance of  both KIN and READ, particularly at lower coverages. This is likely because ROH actually reduces the variance in differences, and so it becomes easier to estimate the location of IBD tracts. For READ, the presence of ROH causes a bias towards inferring closer relatedness classes, but the model we use in KIN reduces this bias  (see Fig. \ref{fig3:Comparison_READ_KIN}).

Contamination can affect the accuracy of relatedness inference for two reasons. For one, if a substantial fraction of samples is contaminated, then the estimation of $p_0$ becomes inaccurate, because the majority of pairwise differences will include at least one contaminated sample. The second issue is that contaminated samples will look less similar to other individuals, and thus cause a bias towards inferring them to be less related to other individuals. The amount of contamination does not need to be large for this to be important, in our simulations we find that even at contamination levels $\leq 3\%$, the performance of READ is substantially reduced.  Assuming the contamination rates are correctly inferred, the correction we implemented leads to improved performance compared to naive methods such as READ, although they too could be ameliorated in a similar way (Laurits et al., 2022 (in press)). SNP ascertainment can cause a slight decrease in power for both  methods, since it reduces the amount of data available. However, the decrease in power is lower for KIN compared to READ.
